---
title: "Demographic Questions"
output: html_document
date: "2023-11-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***Load packages***
```{r}
pacman::p_load(tidyverse, 
							 ggplot2,
							 janitor,
							 lubridate,
							 dplyr,
							 fitdistrplus)
```

### Load data
```{r}
raw_data <- readRDS('/Users/bevinmanuelpillai/Desktop/EPI 569/RFF/rff_finalcases_clean_2023.rds') %>% janitor::clean_names()
roster <- readRDS('/Users/bevinmanuelpillai/Desktop/EPI 569/RFF/rosters_569_517_clean_2023.rds') %>%  janitor::clean_names()
```

### Data Cleaning 
```{r}
linelist <- raw_data %>% 
	#rename variables 
	rename(c('report' = 'reporting_time',
					 'infector_id' = 'infectedby',
					 'exposure_date' = 'date_of_exposure',
					 'exposure_time' = 'time_of_exposure',
					 'symptomatic' = 'did_you_have_symptoms',
					 'incubation_period' = 'how_many_hours_after_exposure_did_you_develop_symptoms',
					 'illness_duration' = 'how_many_hours_after_your_symptom_onset_did_you_feel_better',
					 'n_contacts' = 'how_many_people_did_you_expose',
					 'n_infect' = 'how_many_people_did_you_actually_infect',
					 'contact1_id' = 'case_id_first_exposure',
					 'contact1_date' = 'date_of_first_exposure',
					 'contact1_time' = 'time_of_first_exposure',
					 'contact2_id' = 'case_id_second_exposure',
					 'contact2_date' = 'date_of_second_exposure',
					 'contact2_time' = 'time_of_second_exposure',
					 'contact3_id' = 'case_id_third_exposure',
					 'contact3_date' = 'date_of_third_exposure',
					 'contact3_time' = 'time_of_third_exposure',
					 'contact4_id' = 'case_id_fourth_exposure',
					 'contact4_date' = 'date_of_fourth_exposure',
					 'contact4_time' = 'time_of_fourth_exposure',
					 'contact5_id' = 'case_id_fifth_exposure',
					 'contact5_date' = 'date_of_fifth_exposure',
					 'contact5_time' = 'time_of_fifth_exposure',
					 'exposure0' = 'date_time_exposure',
					 'exposure1' = 'date_time_first_exposure',
					 'exposure2' = 'date_time_second_exposure',
					 'exposure3' = 'date_time_third_exposure',
					 'exposure4' = 'date_time_fourth_exposure',
					 'exposure5' = 'date_time_fifth_exposure')) %>% 
	
	#recode gender variable as numeric (0 = Female, 1 = Male, 2 = Non-binary)
	mutate( gender = recode(gender, "Female" = 0, "Male" = 1, "Non-binary" = 2)) %>% 
	
	#recode course variable as numeric (0 = EPI 569, 1 = EPI 517, 2 = both)
	mutate(course = recode(course, 
                       "EPI 569 (Concepts and Methods in ID EPI)" = 0, 
                       "EPI/GH 517 (Case Studies in ID)" = 1,
                       .default = 2)) %>%

	
	#recode symptomatic variable as numeric (0 = No, 1 = Yes)
	mutate(symptomatic = recode(symptomatic, "No" = 0, "Yes" = 1)) %>% 

	#recode severe variable as numeric (0= No, 1 = Yes)

	mutate(severe = recode(severe, "No" = 0, "Yes" =1))
```

Merge roster and linelist
```{r}
full_data <- left_join(linelist, roster, by = "case_id")
```

Subset to Severe and Non-severe RFF
```{r}
Severe <- linelist %>% 
	filter(severe == 1)

Non_Severe <- linelist %>% 
	filter(severe == 0)
```
Severity by Demographic Factors
```{r}
#Severity by Gender
table(linelist$severe, linelist$gender)

#Severity by Sympomatic vs. asympomatic
table(linelist$severe, linelist$symptomatic)

#Severity by Course
table(linelist$severe, linelist$course)


```
Population Sizes
```{r}
table(roster$course)
#EPI_569 = 54, EPI 517 = 95, EPI 569 + EPI 517 = 5

table(roster$degree)
#Doctor of Philosphy = 4, Liberal Arts & Sciences = 1, Master of Public Health = 136, Master of Science Public Health = 10, Non-Degree Public Health = 3

table(roster$program)
#Behavioral Sciences = 3, Biostatistics = 1, Environmental Health + Epidemiology = 4, Environmental Health = 3, Epidemiology = 75, Global Environmental Health = 6, Global Epidemiology = 29, Global Health = 29, Human Health BA = 1, Special Standing Student = 3.

table(roster$level)
#full standing = 4, Graduate Year 1 = 63, Graduate Year 2 = 83, Non-Degree Graduate = 3, Senior = 1
```
Case Distribution
```{r}
table(full_data$course.y)
#EPI 569 = 28 cases, EPI 517 = 49 cases, Both = 4

table(linelist$gender)
#Female = 49 cases, Male = 17 cases, Non-Binary = 2

table(Severe$gender)
#Females = 20 severe cases, Males = 8 severe cases
```
Illness Duration in severe cases
```{r}
mean(Severe$illness_duration)
sd(Severe$illness_duration)

```

Illness Duration in non_severe cases
```{r}
mean(Non_Severe$illness_duration)
sd(Non_Severe$illness_duration)
```
Compare infectiousness of symptomatic vs asymptomatic cases
```{r}
symptomatic <- linelist %>% 
	filter(symptomatic == 1)

asymptomatic <- linelist %>% 
	filter(symptomatic == 0)

#presenting both total number infected and infected vs number of contacts
#Symptomatic average number of infected
symptomatic %>% 
	summarize(mean(n_infect)) 

#Symptomatic ratio of infected to contacts
symptomatic %>% 
  filter(!is.na(n_infect) & !is.na(n_contacts) & n_contacts != 0) %>%
  mutate(infectiousness = n_infect/n_contacts) %>%
  summarize(mean(infectiousness, na.rm = TRUE))

#asymptomatic average number of infected
asymptomatic %>% 
	summarize(mean(n_infect))

#Asymptomatic ratio of infected to contacts
asymptomatic %>% 
  filter(!is.na(n_infect) & !is.na(n_contacts) & n_contacts != 0) %>%
  mutate(infectiousness = n_infect/n_contacts) %>%
  summarize(mean(infectiousness, na.rm = TRUE))
```
Compare Period Prevalence of Undergraduates and Graduate Year 1 vs those that are Graduate Year 2 and doctorate students
```{r}
#Create new variable to dicotomize level variable
#Get numerator
full_data <- full_data %>% 
	mutate(education = ifelse(level %in% c("Graduate Year 1", "Senior"),0,1))

table(full_data$education)

# 44 = 0 and 37 = 1
#Get Denominator
roster <- roster %>% 
	mutate(education = ifelse(level %in% c("Graduate Year 1", "Senior"),0,1))

table(roster$education)
# 64 = 0 and 90 = 1

period_prevalance_G1S = 44/64
period_prevalance_G2D = 37/90
```

Compare Period Prevalence of RFF for different PH programs
```{r}
table(roster$program)

roster <- roster %>% 
  mutate(program_num = ifelse(program == "Epidemiology", 0,
                              ifelse(program == "Global Epidemiology", 1, 
                                     ifelse(program == "Global Health", 2, 3))))

table(roster$program_num)

full_data <- full_data %>% 
	mutate(program_num = ifelse(program == "Epidemiology", 0,
                              ifelse(program == "Global Epidemiology", 1, 
                                     ifelse(program == "Global Health", 2, 3))))
table(full_data$program_num)

period_prevalance_epi = 38/75
period_prevalance_glepi = 18/29
period_prevalance_gh = 12/29
period_prevalance_otherprs = 13/21
```




