---
title: "EPI 569 Final Project"
author: "Meghna Ray, Charlotte Doran, and Bevin Manuelpillai"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Load packages
pacman::p_load(tidyverse, 
               ggplot2,
               janitor,
               lubridate,
               dplyr,
							 splitstackshape,
							 earlyR,
							 incidence,
							 fitdistrplus,
							 EpiEstim,
							 ggpubr,
							 deSolve,
							 reshape2)
```

### Load data

```{r}
raw_data <- readRDS('rff_finalcases_clean_2023_2.rds') %>% janitor::clean_names()
roster <- readRDS('rosters_569_517_clean_2023_2.rds') %>%  janitor::clean_names()
```

### Data Cleaning

```{r}
linelist <- raw_data %>% 
  #rename variables 
  rename(c('report' = 'reporting_time',
           'infector_id' = 'infectedby',
           'exposure_date' = 'date_of_exposure',
           'exposure_time' = 'time_of_exposure',
           'symptomatic' = 'did_you_have_symptoms',
           'incubation_period' = 'how_many_hours_after_exposure_did_you_develop_symptoms',
           'illness_duration' = 'how_many_hours_after_your_symptom_onset_did_you_feel_better',
           'n_contacts' = 'how_many_people_did_you_expose',
           'n_infect' = 'how_many_people_did_you_actually_infect',
           'contact1_id' = 'case_id_first_exposure',
           'contact1_date' = 'date_of_first_exposure',
           'contact1_time' = 'time_of_first_exposure',
           'contact2_id' = 'case_id_second_exposure',
           'contact2_date' = 'date_of_second_exposure',
           'contact2_time' = 'time_of_second_exposure',
           'contact3_id' = 'case_id_third_exposure',
           'contact3_date' = 'date_of_third_exposure',
           'contact3_time' = 'time_of_third_exposure',
           'contact4_id' = 'case_id_fourth_exposure',
           'contact4_date' = 'date_of_fourth_exposure',
           'contact4_time' = 'time_of_fourth_exposure',
           'contact5_id' = 'case_id_fifth_exposure',
           'contact5_date' = 'date_of_fifth_exposure',
           'contact5_time' = 'time_of_fifth_exposure',
           'exposure0' = 'date_time_exposure',
           'exposure1' = 'date_time_first_exposure',
           'exposure2' = 'date_time_second_exposure',
           'exposure3' = 'date_time_third_exposure',
           'exposure4' = 'date_time_fourth_exposure',
           'exposure5' = 'date_time_fifth_exposure',
  				 'severe_date' = 'date_you_became_severe')) %>% 
  
  #recode gender variable as numeric (0 = Female, 1 = Male, 2 = Non-binary)
  mutate(gender = recode(gender, "Female" = 0, "Male" = 1, "Non-binary" = 2)) %>% 
  
  #recode course variable as numeric (0 = EPI 569, 1 = EPI 517, 2 = both)
  mutate(course = recode(course, "EPI 569 (Concepts and Methods in ID EPI)" = 0, 
                         "EPI/GH 517 (Case Studies in ID)" = 1,
                         "BOTH" = 2)) %>% 
  
  #recode symptomatic variable as numeric (0 = No, 1 = Yes)
  mutate(symptomatic = recode(symptomatic, "No" = 0, "Yes" = 1)) %>% 
	
	#recode severe variable as numeric (0 = No, 1 = Yes)
	mutate(severe = recode(severe, "No" = 0, "Yes" = 1))
  
```

### Important values

```{r}
total_pop <- nrow(roster)
total_cases <- nrow(linelist)
palette <- c("#66B4BE", "#7BA536", "#E28571", "#636467", "#49494B", "#1C3052", "#24525A", "#548135", "#0D3840", "#722C21")
names(palette) <- c("blue",
										"green",
										"red",
										"gray",
										"dark_gray",
										"dark_blue",
										"gray_blue",
										"dark_green",
										"dark_gray_blue",
										"dark_red")
```

# Table 1

```{r}
roster %>% 
	group_by(course) %>% 
	summarize(count = n(),
						percent = (count/total_pop) * 100)

roster %>% 
	group_by(program) %>% 
	summarize(count = n(),
						percent = (count/total_pop) * 100)

linelist %>% 
	group_by(symptomatic, severe) %>% 
	summarize(count = n(),
						percent = n()/total_cases * 100)

linelist %>% 
	filter(symptomatic == 1) %>% 
	group_by(severe) %>% 
	summarize(count = n(),
						percent = count/sum(linelist$symptomatic, na.rm = TRUE) * 100)

linelist_roster <- linelist %>% 
	left_join(roster, by = c("case_id" = "case_id"))

linelist_roster %>% 
		group_by(course.y) %>% 
		summarize(count = n(),
							percent = (count/total_cases) * 100)

```

# Plot Epi Curve

## Onset and recovery

```{r}
med_incubation <- median(linelist$incubation_period, na.rm = TRUE)
med_duration <- median(linelist$illness_duration, na.rm = TRUE)
asymp_duration <- med_incubation + med_duration

#Create onset variable
linelist <- linelist %>% 
	mutate(onset = case_when(symptomatic == 0 ~ NA, #no symptom onset for asymptomatic cases
													 symptomatic == 1 ~ exposure0 + hours(incubation_period), #symptoms start after incubation period
													 is.na(symptomatic) ~ NA, #keep missing and NA's NA
													 TRUE ~ NA)) %>% 
  mutate(onset = date(onset))

#Create recovered variable
linelist <- linelist %>% 
	mutate(recovered = case_when((symptomatic == 0 | is.na(symptomatic)) ~ exposure0 + hours(asymp_duration), #imputed infectious periods for asymptomatic or missing
															 (symptomatic == 1 & severe == 1) ~ pmax(onset + hours(illness_duration), severe_date), #for severe cases, recovery is the later of illness duration end or severe date
															 (symptomatic == 1 & severe == 0) ~ onset + hours(illness_duration),
															 TRUE ~ NA)) %>% 
	mutate(recovered = date(recovered))
```

## Epi curve by onset date

```{r}
#Create data frame for epi curve with onset and incidence
epicurve1 <- as.data.frame(linelist %>% 
                            group_by(onset) %>%
                            summarize(incidence=n()))
#Plot epi curve 
plot1 <- epicurve1 %>% 
	ggplot(aes(x = onset, y = incidence, labs = FALSE)) + 
  geom_bar(stat="identity", fill=palette["dark_gray_blue"]) +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous(breaks = seq(0, 15, by = 2)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
  			panel.grid = element_blank()) +
  labs(x = "Onset Date",
       y = "Number of Cases",
       title = "Epi Curve for Rollins Fall Fever")
plot1
```

## Epi curve by exposure date

```{r}
epicurve2 <- as.data.frame(linelist %>% 
                            group_by(exposure_date) %>%
                            summarize(incidence=n()))

#Plot epi curve 
plot2 <- epicurve2 %>% 
  ggplot(aes(x = exposure_date, y = incidence, labs = FALSE)) + 
  geom_bar(stat="identity", fill=palette["dark_gray_blue"]) +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous(breaks = seq(0, 15, by = 2)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
  			panel.grid = element_blank()) +
  labs(x = "Exposure Date",
       y = "Number of Cases",
       title = "Epi Curve for Rollins Fall Fever")
plot2
```

## Log cases

```{r, include = TRUE, echo = T}
epicurve3 <- epicurve2 %>% 
  mutate(log_cases = log(incidence))

#Plot log cases
plot2 <- epicurve3 %>% 
  ggplot(aes(x = exposure_date, y = log_cases, labs = FALSE)) + 
  geom_bar(stat="identity", fill=palette["dark_gray_blue"]) +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
  			panel.grid = element_blank()) +
  labs(x = "Exposure Date",
       y = "Log Cases",
       title = "Epi Curve - Log Cases")
plot2
```

## Susceptible, infected, recovered by day

***Calculate number of recovered by day***

```{r}
recovered <- linelist %>% 
  group_by(recovered) %>% 
  summarize(n_recovered = n()) %>% 
  na.omit()
```

***Calculate number of infected by day***

```{r}
infected <- linelist %>%
  group_by(exposure_date) %>%
  summarize(n_exposed = n()) %>%
  na.omit()

#Create a sequence of dates from the onset to the end of the outbreak
dateseq <- seq(from = as.Date(min(linelist$exposure_date, na.rm = TRUE)),
								 to = as.Date(max(linelist$recovered, na.rm = TRUE)),
								 by = "1 day")

#For each date, calculate the number of infected people by determining if that date is between their onset date and their recovery date
infectedseq <- map(dateseq, ~sum(.x >= linelist$exposure_date & .x < linelist$recovered, na.rm = TRUE))
infected2 <- data.frame(date = dateseq, n_infected = as.numeric(infectedseq))
```

***Calculate number of susceptible by day***

```{r}
#Join infected and recovered datasets
epicurve4 <- infected2 %>% 
  full_join(recovered, by = c("date" = "recovered")) %>% 
	full_join(infected, by = c("date" = "exposure_date")) %>% 
	#Set na values to zero 
  mutate(n_infected = ifelse(is.na(n_infected), 0, n_infected)) %>% 
  mutate(n_recovered = ifelse(is.na(n_recovered), 0, n_recovered)) %>% 
	mutate(n_exposed = ifelse(is.na(n_exposed), 0, n_exposed)) %>% 
	#Create variable for cumulative recoveries and infected
  mutate(total_recovered = cumsum(n_recovered)) %>% 
  mutate(total_infected = cumsum(n_exposed))

#Calculate number of people susceptible 
epicurve4 <- epicurve4 %>% 
  mutate(n_susceptible = total_pop - total_infected) %>% 
	#Check that the total population size is stable
	mutate(total = n_infected + total_recovered + n_susceptible)
```

***Plot susceptible, infected, and recovered by day***

```{r}
plot3 <- ggplot(epicurve4, aes(x = date)) +
  geom_line(aes(y = n_susceptible, color = "Susceptible"), linewidth = 2) +
  geom_line(aes(y = n_infected, color = "Infected"), linewidth = 2) +
  geom_line(aes(y = total_recovered, color = "Recovered"), linewidth = 2) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Date",
       y = "Number of Cases",
       title = "Epi Curve for Rollins Fall Fever") +
  scale_color_manual(name = "Class",
                     breaks = c("Susceptible", "Infected", "Recovered"),
                     values = c("Susceptible" = as.character(palette["blue"]),
                     				    "Infected" = as.character(palette["green"]),
                     				    "Recovered" = as.character(palette["red"]))) +
	scale_x_date(breaks = "2 days",
							 date_labels = "%b %d",
							 expand = c(0.05,0,0.15, 0.15)) +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 8),
  			legend.key = element_rect(fill = "white"),
  			panel.grid = element_blank())

plot3
```

**Plot epi curve using only symptom onset dates for observed (symptomatic) cases**

```{r}
epicurve5 <- as.data.frame(linelist %>% 
													 	filter(symptomatic == 1) %>% 
                            group_by(onset) %>%
                            summarize(incidence=n()))

#Plot epi curve 
plot5 <- epicurve5 %>% 
  ggplot(aes(x = onset, y = incidence, labs = FALSE)) + 
  geom_bar(stat="identity", fill=palette["dark_gray_blue"]) +
  scale_x_date(date_breaks = "2 days", date_labels = "%b %d") +
  scale_y_continuous(breaks = seq(0, 15, by = 2)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
  			panel.grid = element_blank()) +
  labs(x = "Symptom Onset Date",
       y = "Number of Cases",
       title = "Epi Curve by Onset Date for Symptomatic Cases of Rollins Fall Fever")
plot5
```

# Natural History

## Incubation period

```{r}
mean_incubation <- linelist %>%
	summarise(mean(incubation_period, na.rm = T))
#23.36

sd_incubation <- linelist %>% 
	summarise(sd(incubation_period, na.rm = T))
#7.38
```

## Illness duration

```{r}
mean_duration <- linelist %>%
	summarise(mean(illness_duration, na.rm = T))
#25.02

ds_duration <- linelist %>%
	summarise(sd(illness_duration, na.rm = T))
#9.09
```

## Serial interval

```{r}
#Select only the columns of interest
serial <- raw_data[, c("case_id", "infectedby", "date_of_exposure", "time_of_exposure", "how_many_hours_after_exposure_did_you_develop_symptoms")]

#Covert to Exposure to Date_Time format
serial$date_time_exposure <- as.POSIXct(paste(serial$date_of_exposure, serial$time_of_exposure), format="%Y-%m-%d %H:%M:%S")
serial <- serial[, !(names(serial) %in% c("date_of_exposure", "time_of_exposure"))]

#Covert Symptom Onset to Date_Time Format
serial$date_time_symptom_onset <- serial$date_time_exposure + 
  as.difftime(serial$how_many_hours_after_exposure_did_you_develop_symptoms, units = "hours")

#Remove incubation period column
serial <- serial[, !(names(serial) %in% c("how_many_hours_after_exposure_did_you_develop_symptoms"))]

#Get infector and infected columns
merged_data <- merge(serial, serial, by.x = "infectedby", by.y = "case_id", suffixes = c("_infected", "_infector"))

#Calculate serial interval in days
merged_data$serial_interval_in_days <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "days"))

#Calculate serial interval in hours
merged_data$serial_interval_in_hours <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "hours"))

#Calculate serial interval in minutes
merged_data$serial_interval_in_mins <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "mins"))

serial_interval_in_days_average<- mean(merged_data$serial_interval_in_days, na.rm = T)

serial_interval_in_days_sd<- sd(merged_data$serial_interval_in_days, na.rm = T)

serial_interval_in_hours_average<- mean(merged_data$serial_interval_in_hours, na.rm = T)

serial_interval_in_hours_sd<- sd(merged_data$serial_interval_in_hours, na.rm = T)
```

## Latent Period

```{r}
serial$latent <- serial$date_time_symptom_onset - serial$date_time_exposure
mean_latent <- mean(serial$latent, na.rm = TRUE)
#23.36
ds_latent <- sd(serial$latent, na.rm = TRUE)
#7.38
```

# Reproduction Numbers

```{r}
start_day <- date("2023-10-11")
end_day <- date("2023-10-20")
```

We used the initial growth rate method to calculate $R_0$ from the plot of the log cases based on exposure date. The first case on RFF was observed on `r start_day`. Using the plot of log cases, the log-linear growth period ends on `r end_day`. $I(t) \approx I(0) e^{rt}$

```{r}
linear_growth <- epicurve3 %>% 
	filter(exposure_date %in% c(start_day, end_day))

plot6 <- plot2 +
	geom_line(data = linear_growth,
						mapping = aes(x = exposure_date, y = log_cases),
						color = palette["green"],
						linewidth = 1.5,
						linetype = "dashed")
plot6
```

```{r}
r0_df <- epicurve3 %>%
	filter(exposure_date >= start_day, exposure_date <= end_day) %>% 
	splitstackshape::expandRows("incidence", drop = FALSE) %>% 
	dplyr::select(exposure_date) %>% 
	pull(exposure_date)

i <- incidence(r0_df)
mu <- serial_interval_in_days_average #replace with mean generation time
sigma <- serial_interval_in_days_sd #replace with standard deviation of generation time
res <- get_R(i, si_mean = mu, si_sd = sigma)
R_val <- sample_R(res, 1000)
r0 <- mean(R_val)
r0
#~2.42
```

Based on the initial growth rate, the basic reproductive number for RFF, $R_0$, is `r round(r0, 1)`.

We can compare this value to the $R_0$ calculated based purely on the number of people that each individual exposed.

```{r}
r0_2 <- mean(linelist$n_contacts, na.rm = TRUE)
#2.04

#By symptom status
r0_3 <- linelist %>% 
	group_by(symptomatic) %>% 
	summarize(r0 = mean(n_contacts, na.rm = TRUE),
						count = n())

#By severity
r0_4 <- linelist %>% 
	group_by(severe) %>% 
	summarize(r0 = mean(n_contacts, na.rm = TRUE),
						count = n())
```

# Modeling

## SEIR Equations

$\frac{dS}{dt} = -\lambda_t S$

$\frac{dE}{dt} = \lambda_t S - \gamma E$

$\frac{dI}{dt} = \gamma E - \sigma I$

$\frac{dR}{dt} = \sigma I$

## Waning Immunity Equations

$\frac{dS}{dt} = \omega R -\lambda_t S$

$\frac{dE}{dt} = \lambda_t S - \gamma E$

$\frac{dI}{dt} = \gamma E - 0.65 \sigma I - 0.35 \mu I$

$\frac{dR}{dt} = 0.65 \sigma I - \omega R$

$\frac{dR_s}{dt} = 0.35 \mu I$

## Observed Data

```{r}
mean_duration_severe <- linelist %>%
	filter(severe == 1) %>%
	summarize(mean = mean(illness_duration, na.rm = TRUE)) %>% 
	as.numeric()
#24.39286

mean_duration_nonsevere <- linelist %>%
	filter(severe == 0) %>%
	summarize(mean = mean(illness_duration, na.rm = TRUE)) %>% 
	as.numeric()
#25.66667

mean_latent <- as.numeric(mean_latent)
#23.36364



```

```{r}
#Create function for model equations
sir_ode <- function(times,init,parms){
  with(as.list(c(parms,init)), {
    # ODEs
    dS <- (omega*R) - (alpha*beta*I*S/(S+E+I+R))
    dE <- (alpha*beta*I*S/(S+E+I+R)) - (gamma*E)
    dI <- (gamma*E) - (0.65*sigma*I) - (0.35*mu*I)
    dR <- (0.65*sigma*I)  - (omega*R)
    dRs <- 0.35*mu*I
    list(c(dS,dE,dI,dR,dRs))
  })
}

#Define parameters
parms <- c(alpha = 5,              # alpha = daily contacts
           beta = 0.26,             # beta = probability of infection on contact
				 	 gamma = 1/(mean_latent/24),  # gamma = rate of infectiousness onset
           sigma = 1/(mean_duration_nonsevere/24),# sigma = rate of recovery per day
           mu = 1/(mean_duration_severe/24),               # mu = rate of severe disease
           omega = 0)            # omega = rate of immune loss per day

#Set initial conditions
init <- c(S = 153,
					E = 0,
					I = 1,
					R = 0,
					R_s = 0)

times <- seq(0, 30, length.out=30)
sir_out <- lsoda(init,times,sir_ode,parms)
sir_out_long <- melt(as.data.frame(sir_out),"time")

#Plot the model output
sir_out_long %>% ggplot(aes(x=time, y=value, colour=variable, group=variable)) +
	geom_line(lwd = 2) +
  labs(x = "Time since index case exposure (days)",
  		 y = "Number of cases",
  		 title = "SEIR Model of RFF") +
	scale_color_manual(name = "Class",
										 breaks = c("S", "E", "I", "R", "R_s"),
										 labels = c("Susceptible",
										 					"Exposed",
										 					"Infectious",
										 					"Recovered",
										 					"Became severe"),
										 values = c("S" = as.character(palette["blue"]),
										 					 "E" = as.character(palette["dark_green"]),
										 					 "I" = as.character(palette["green"]),
										 					 "R" = as.character(palette["red"]),
										 					 "R_s" = as.character(palette["dark_red"]))) +
	theme(panel.grid = element_blank())

as.numeric(sir_out_long %>% 
	filter(variable == "R") %>% 
	summarize(max(value))) +
	as.numeric(sir_out_long %>% 
	filter(variable == "R_s") %>% 
	summarize(max(value)))
```

## Waning Immunity

```{r}
mean_duration_severe <- linelist %>%
	filter(severe == 1) %>%
	summarize(mean = mean(illness_duration, na.rm = TRUE)) %>% 
	as.numeric()
#24.39286

mean_duration_nonsevere <- linelist %>%
	filter(severe == 0) %>%
	summarize(mean = mean(illness_duration, na.rm = TRUE)) %>% 
	as.numeric()
#25.66667

mean_latent <- as.numeric(mean_latent)
#23.36364



```

```{r}
#Create function for model equations
sir_ode <- function(times,init,parms){
  with(as.list(c(parms,init)), {
    # ODEs
    dS <- (omega*R) - (alpha*beta*I*S/(S+E+I+R))
    dE <- (alpha*beta*I*S/(S+E+I+R)) - (gamma*E)
    dI <- (gamma*E) - (0.65*sigma*I) - (0.35*mu*I)
    dR <- (0.65*sigma*I)  - (omega*R)
    dRs <- 0.35*mu*I
    list(c(dS,dE,dI,dR,dRs))
  })
}

#Define parameters
parms <- c(alpha = 5,              # alpha = daily contacts
           beta = 0.26,             # beta = probability of infection on contact
				 	 gamma = 1/(mean_latent/24),  # gamma = rate of infectiousness onset
           sigma = 1/(mean_duration_nonsevere/24),# sigma = rate of recovery per day
           mu = 1/(mean_duration_severe/24),               # mu = rate of severe disease
           omega = 0)            # omega = rate of immune loss per day

#Set initial conditions
init <- c(S = 153,
					E = 0,
					I = 1,
					R = 0,
					R_s = 0)

times <- seq(0, 30, length.out=30)
sir_out <- lsoda(init,times,sir_ode,parms)
sir_out_long <- melt(as.data.frame(sir_out),"time")

#Plot the model output
sir_out_long %>% ggplot(aes(x=time, y=value, colour=variable, group=variable)) +
	geom_line(lwd = 2) +
  labs(x = "Time since index case exposure (days)",
  		 y = "Number of cases",
  		 title = "SEIR Model of RFF") +
	scale_color_manual(name = "Class",
										 breaks = c("S", "E", "I", "R", "R_s"),
										 labels = c("Susceptible",
										 					"Exposed",
										 					"Infectious",
										 					"Recovered",
										 					"Became severe"),
										 values = c("S" = as.character(palette["blue"]),
										 					 "E" = as.character(palette["dark_green"]),
										 					 "I" = as.character(palette["green"]),
										 					 "R" = as.character(palette["red"]),
										 					 "R_s" = as.character(palette["dark_red"]))) +
	theme(panel.grid = element_blank())

as.numeric(sir_out_long %>% 
	filter(variable == "R") %>% 
	summarize(max(value))) +
	as.numeric(sir_out_long %>% 
	filter(variable == "R_s") %>% 
	summarize(max(value)))
```
