---
title: "EPI 569 Final Project"
author: "Meghna Ray, Charlotte Doran, and Bevin Manuelpillai"
date: "`r Sys.Date()`"
output: html_document
execute:
	cache: true
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Load packages
pacman::p_load(ggplot2,
               janitor,
               lubridate,
							 splitstackshape,
							 earlyR,
							 incidence,
							 fitdistrplus,
							 EpiEstim,
							 ggpubr,
							 deSolve,
							 reshape2,
							 tidyverse)
```

### Load data

```{r}
raw_data <- readRDS('rff_finalcases_clean_2023_2.rds') %>% janitor::clean_names()
roster <- readRDS('rosters_569_517_clean_2023_2.rds') %>%  janitor::clean_names()
```

### Data Cleaning

```{r}
linelist <- raw_data %>% 
  #rename variables 
  rename(c('report' = 'reporting_time',
           'infector_id' = 'infectedby',
           'exposure_date' = 'date_of_exposure',
           'exposure_time' = 'time_of_exposure',
           'symptomatic' = 'did_you_have_symptoms',
           'incubation_period' = 'how_many_hours_after_exposure_did_you_develop_symptoms',
           'illness_duration' = 'how_many_hours_after_your_symptom_onset_did_you_feel_better',
           'n_contacts' = 'how_many_people_did_you_expose',
           'n_infect' = 'how_many_people_did_you_actually_infect',
           'contact1_id' = 'case_id_first_exposure',
           'contact1_date' = 'date_of_first_exposure',
           'contact1_time' = 'time_of_first_exposure',
           'contact2_id' = 'case_id_second_exposure',
           'contact2_date' = 'date_of_second_exposure',
           'contact2_time' = 'time_of_second_exposure',
           'contact3_id' = 'case_id_third_exposure',
           'contact3_date' = 'date_of_third_exposure',
           'contact3_time' = 'time_of_third_exposure',
           'contact4_id' = 'case_id_fourth_exposure',
           'contact4_date' = 'date_of_fourth_exposure',
           'contact4_time' = 'time_of_fourth_exposure',
           'contact5_id' = 'case_id_fifth_exposure',
           'contact5_date' = 'date_of_fifth_exposure',
           'contact5_time' = 'time_of_fifth_exposure',
           'exposure0' = 'date_time_exposure',
           'exposure1' = 'date_time_first_exposure',
           'exposure2' = 'date_time_second_exposure',
           'exposure3' = 'date_time_third_exposure',
           'exposure4' = 'date_time_fourth_exposure',
           'exposure5' = 'date_time_fifth_exposure',
  				 'severe_date' = 'date_you_became_severe')) %>% 
	mutate(exposure1 = as_datetime(exposure1),
				 exposure2 = as_datetime(exposure2),
				 exposure3 = as_datetime(exposure3),
				 exposure4 = as_datetime(exposure4),
				 exposure5 = as_datetime(exposure4)) %>% 
  
  #recode gender variable as numeric (0 = Female, 1 = Male, 2 = Non-binary)
  mutate(gender = recode(gender, "Female" = 0, "Male" = 1, "Non-binary" = 2)) %>% 
  
  #recode course variable as numeric (0 = EPI 569, 1 = EPI 517, 2 = both)
  mutate(course = recode(course, "EPI 569 (Concepts and Methods in ID EPI)" = 0, 
                         "EPI/GH 517 (Case Studies in ID)" = 1,
                         "BOTH" = 2)) %>% 
  
  #recode symptomatic variable as numeric (0 = No, 1 = Yes)
  mutate(symptomatic = recode(symptomatic, "No" = 0, "Yes" = 1)) %>% 
	
	#recode severe variable as numeric (0 = No, 1 = Yes)
	mutate(severe = recode(severe, "No" = 0, "Yes" = 1))
  
```

### Important values
```{r}
total_pop <- nrow(roster)
palette <- c("#66B4BE", "#7BA536", "#E28571", "#636467")
```


# Plot Epi Curve

## Onset and recovery

```{r}
#Create new onset variable using date and time of exposure + incubation period 
med_incubation <- median(linelist$incubation_period, na.rm = TRUE)
med_duration <- median(linelist$illness_duration, na.rm = TRUE)
asymp_duration <- med_incubation + med_duration

#Create onset variable
linelist <- linelist %>% 
	mutate(onset_datetime = case_when(symptomatic == 0 ~ NA, #no symptom onset for asymptomatic cases
													 symptomatic == 1 ~ exposure0 + hours(incubation_period), #symptoms start after incubation period
													 is.na(symptomatic) ~ NA, #keep missing and NA's NA
													 TRUE ~ NA)) %>% 
  mutate(onset = date(onset_datetime))

#Create recovered variable
linelist <- linelist %>% 
	mutate(recovered_datetime = case_when((symptomatic == 0 | is.na(symptomatic)) ~ exposure0 + hours(asymp_duration), #imputed infectious periods for asymptomatic or missing
															 (symptomatic == 1 & severe == 1) ~ pmax(onset + hours(illness_duration), severe_date), #for severe cases, recovery is the later of illness duration end or severe date
															 (symptomatic == 1 & severe == 0) ~ onset + hours(illness_duration),
															 TRUE ~ NA)) %>% 
	mutate(recovered = date(recovered_datetime))
```

## Epi curve by onset date

```{r}
#Create data frame for epi curve with onset and incidence
epicurve1 <- as.data.frame(linelist %>% 
                            group_by(onset) %>%
                            summarize(incidence=n()))
#Plot epi curve 
plot1 <- epicurve1 %>% 
  ggplot(aes(x = onset, y = incidence, labs = FALSE)) + 
  geom_bar(stat="identity", fill=palette[4]) +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous(breaks = seq(0, 15, by = 2)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Onset Date",
       y = "Number of Cases",
       title = "Epi Curve for Rollins Fall Fever")
plot1
```

## Epi curve by exposure date

```{r}
epicurve2 <- as.data.frame(linelist %>% 
                            group_by(exposure_date) %>%
                            summarize(incidence=n()))

#Plot epi curve 
plot4 <- epicurve2 %>% 
  ggplot(aes(x = exposure_date, y = incidence, labs = FALSE)) + 
  geom_bar(stat="identity", fill=palette[4]) +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous(breaks = seq(0, 15, by = 2)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Exposure Date",
       y = "Number of Cases",
       title = "Epi Curve for Rollins Fall Fever")
plot4
```

## Log cases

```{r, include = TRUE, echo = T}
epicurve3 <- epicurve2 %>% 
  mutate(log_cases = log(incidence))

#Plot log cases
plot2 <- epicurve3 %>% 
  ggplot(aes(x = exposure_date, y = log_cases, labs = FALSE)) + 
  geom_bar(stat="identity", fill=palette[4]) +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Exposure Date",
       y = "Log Cases",
       title = "Epi Curve - Log Cases")
plot2
```

## Susceptible, infected, recovered by day
***Calculate number of recovered by day***

```{r}
recovered <- linelist %>% 
  group_by(recovered) %>% 
  summarize(n_recovered = n()) %>% 
  na.omit()
```

***Calculate number of infected by day***

```{r}
infected <- linelist %>%
  group_by(exposure_date) %>%
  summarize(n_exposed = n()) %>%
  na.omit()

#Create a sequence of dates from the onset to the end of the outbreak
dateseq <- seq(from = as.Date(min(linelist$exposure_date, na.rm = TRUE)),
								 to = as.Date(max(linelist$recovered, na.rm = TRUE)),
								 by = "1 day")

#For each date, calculate the number of infected people by determining if that date is between their onset date and their recovery date
infectedseq <- map(dateseq, ~sum(.x >= linelist$exposure_date & .x < linelist$recovered, na.rm = TRUE))
infected2 <- data.frame(date = dateseq, n_infected = as.numeric(infectedseq))
```

***Calculate number of susceptible by day***

```{r}
#Join infected and recovered datasets
epicurve4 <- infected2 %>% 
  full_join(recovered, by = c("date" = "recovered")) %>% 
	full_join(infected, by = c("date" = "exposure_date")) %>% 
	#Set na values to zero 
  mutate(n_infected = ifelse(is.na(n_infected), 0, n_infected)) %>% 
  mutate(n_recovered = ifelse(is.na(n_recovered), 0, n_recovered)) %>% 
	mutate(n_exposed = ifelse(is.na(n_exposed), 0, n_exposed)) %>% 
	#Create variable for cumulative recoveries and infected
  mutate(total_recovered = cumsum(n_recovered)) %>% 
  mutate(total_infected = cumsum(n_exposed))

#Calculate number of people susceptible 
epicurve4 <- epicurve4 %>% 
  mutate(n_susceptible = total_pop - total_infected) %>% 
	#Check that the total population size is stable
	mutate(total = n_infected + total_recovered + n_susceptible)
```

***Plot susceptible, infected, and recovered by day***

```{r}
plot3 <- ggplot(epicurve4, aes(x = date)) +
  geom_line(aes(y = n_susceptible, color = "Susceptible"), linewidth = 1) +
  geom_line(aes(y = n_infected, color = "Infected"), linewidth = 1) +
  geom_line(aes(y = total_recovered, color = "Recovered"), linewidth = 1) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Date",
       y = "Number of Cases",
       title = "Epi Curve for Rollins Fall Fever") +
  scale_color_manual(name='Class',
                     breaks=c('Susceptible', 'Infected', 'Recovered'),
                     values=c('Susceptible'=palette[1], 'Infected'=palette[2], 'Recovered'=palette[3])) +
	scale_x_date(breaks = "2 days",
							 date_labels = "%b %d") +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 8),
  			panel.grid.minor = element_blank())
plot3
```

# Natural History Parameters

**Merge roster and linelist**
```{r}
full_data <- left_join(linelist, roster, by = "case_id")
```

**Subset to Severe and Non-severe RFF**
```{r}
Severe <- full_data %>% 
	filter(severe == 1)

Non_Severe <- full_data %>% 
	filter(severe == 0)
```

**Subset linelist to Severe and Non-severe RFF**
```{r}
Severe_cases <- linelist %>% 
	filter(severe == 1)

Non_Severe_cases <- linelist %>% 
	filter(severe == 0)
```

**Subset to Symptomatic and Asymptomatic**
```{r}
symptomatic <- linelist %>% 
  filter(symptomatic == 1)

asymptomatic <- linelist %>% 
  filter(symptomatic == 0)


nonsevere_symptomatic <- linelist %>% 
  filter(symptomatic == 1 & severe != 1)
```


## Incubation period

```{r}
mean_incubation <- linelist %>%
	summarise(mean(incubation_period, na.rm = T)) %>% 
	as.numeric()
#23.36

sd_incubation <- linelist %>% 
	summarise(sd(incubation_period, na.rm = T)) %>% 
	as.numeric()
#7.38
```


**Incubation period for Severe cases and Non_severe cases**

```{r}
mean_incubation_severe <- mean(Severe_cases$incubation_period)
#23.43

sd_incubation_severe <- sd(Severe_cases$incubation_period)
#6.38

mean_incubation_nonsevere <- mean(Non_Severe_cases$incubation_period)
#23.30

sd_incubation_nonsevere <- sd(Non_Severe_cases$incubation_period)
#8.43

#incubation period
mean(nonsevere_symptomatic$incubation_period)
sd(nonsevere_symptomatic$incubation_period)

#latent period
mean(nonsevere_symptomatic$latent_period)
sd(nonsevere_symptomatic$latent_period)

#duration of illness
mean(nonsevere_symptomatic$illness_duration)
sd(nonsevere_symptomatic$illness_duration)

#infectious period
mean(nonsevere_symptomatic$infectious_period)
sd(nonsevere_symptomatic$infectious_period)
```


**Latent period for asymptomatic cases and symptomatic cases**
```{r}
mean(symptomatic$latent_period)
#18.03
sd(symptomatic$latent_period)
#9.25
mean(asymptomatic$latent_period, na.rm = T)
#15.89
sd(asymptomatic$latent_period, na.rm = T)
#11.08
```


## Illness duration
```{r}
mean_duration <- linelist %>%
	summarise(mean = mean(illness_duration, na.rm = T)) %>% 
	as.numeric()
#25.02

ds_duration <- linelist %>%
	summarise(sd = sd(illness_duration, na.rm = T)) %>% 
	as.numeric()
#9.09
```


## Infectious Period

```{r}
# Convert to datetime
linelist <- linelist %>% 
	mutate(severe_datetime = as_datetime(severe_date))

# #Create datetime Symptom_Onset Variable
# linelist$date_time_symptom_onset <- linelist$exposure0 + 
#   as.difftime(linelist$incubation_period, units = "hours")
# 
# #Create datetime Recovery_date variable
# linelist$date_time_recovery <- linelist$date_time_symptom_onset + 
#   as.difftime(linelist$illness_duration, units = "hours")

#Convert to unambiguous date_time format
# List of variables to convert
# date_time_columns <- c("exposure0", "exposure1", "exposure2", "exposure3", "exposure4", "exposure5", "date_time_symptom_onset", "date_time_recovery", "severe_date")

# #Loop through the list and convert each column
# for (col_name in date_time_columns) {
#   linelist[[col_name]] <- as.POSIXct(linelist[[col_name]], format = "%Y-%m-%d %H:%M:%S")
# }

#Calculate start of infectiousness
linelist <- linelist %>% 
  rowwise() %>% 
  mutate(infectiousness_start = pmin(exposure1, exposure2, exposure3, exposure4, exposure5, onset_datetime, na.rm = TRUE))

#Calculated end of infectiousness
linelist <- linelist %>% 
	rowwise() %>% 
	mutate(infectiousness_end = pmax(exposure1, exposure2, exposure3, exposure4, exposure5, recovered_datetime, na.rm = TRUE))


#Calculate infectious period
linelist$infectious_period <- as.numeric(difftime(linelist$infectiousness_end, linelist$infectiousness_start, units = "hours"))

mean_infectious <- mean(linelist$infectious_period, na.rm = TRUE)
#66.1
sd_infectious <- sd(linelist$infectious_period, na.rm = TRUE)
#43.9

```


## Latent Period

```{r}
linelist$latent_period <- as.numeric(difftime(linelist$infectiousness_start, linelist$exposure0, units = "hours"))

mean_latent <- mean(linelist$latent_period, na.rm = TRUE)
#16.0
sd_latent <- sd(linelist$latent_period, na.rm = TRUE)
#10.5
```

## Mean and SD for Infectious Period Stratified by symptom status

```{r}
# Infectious Period for symptomatic
symptomatic <- linelist %>% 
  filter(symptomatic == 1)
mean_infectious_symptomatic <- mean(symptomatic$infectious_period, na.rm = TRUE)
#69.7
sd_infectious_symptomatic <- sd(symptomatic$infectious_period, na.rm = TRUE)
#45.3

# Infectious Period for asymptomatic
asymptomatic <- linelist %>% 
  filter(symptomatic == 0)
mean_infectious_asymptomatic <- mean(asymptomatic$infectious_period, na.rm = TRUE)
#38.1
sd_infectious_asymptomatic <- sd(asymptomatic$infectious_period, na.rm = TRUE)
#11.1

```

**Subset linelist to Severe and Non-severe RFF**
```{r}
Severe_cases <- linelist %>% 
	filter(severe == 1)

Non_Severe_cases <- linelist %>% 
	filter(severe == 0 & symptomatic == 1)
```

## Natural History Stratified by Severity
```{r}
# Infectious Period for Severe
mean(Severe_cases$infectious_period, na.rm = TRUE)
sd(Severe_cases$infectious_period, na.rm = TRUE)

# Infectious Period for not severe
mean(Non_Severe_cases$infectious_period, na.rm = TRUE)
sd(Non_Severe_cases$infectious_period, na.rm = TRUE)

# Latent Period for Severe
mean(Severe_cases$latent_period, na.rm = TRUE)
sd(Severe_cases$latent_period, na.rm = TRUE)
```

## Serial Interval

**Serial Interval (By symptom onset date)**

```{r}
#Get infector and infected columns
merged_data <- merge(linelist, linelist, by.x = "infector_id", by.y = "case_id", suffixes = c("_infected", "_infector"))

#Calculate serial interval in days
merged_data$serial_interval_in_days <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "days"))

serial_interval_in_days_average <- mean(merged_data$serial_interval_in_days, na.rm = T)
serial_interval_in_days_sd<- sd(merged_data$serial_interval_in_days, na.rm = T)

#Calculate serial interval in hours
merged_data$serial_interval_in_hours <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "hours"))

serial_interval_in_hours_average <- mean(merged_data$serial_interval_in_hours, na.rm = T)
serial_interval_in_hours_sd <- sd(merged_data$serial_interval_in_hours, na.rm = T)

print(serial_interval_in_hours_average)
#43.5
print(serial_interval_in_hours_sd)
#52.2

#Calculate serial interval in minutes
merged_data$serial_interval_in_mins <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "mins"))
```

**Serial Interval (By Infectiousness End Date)**
```{r}
#Get infector and infected columns
merged_data <- merge(linelist, linelist, by.x = "infector_id", by.y = "case_id", suffixes = c("_infected", "_infector"))

#Calculate serial interval in days
merged_data$serial_interval_in_days2 <- as.numeric(difftime(merged_data$infectiousness_end_infected, merged_data$infectiousness_end_infector, units = "days"))

serial_interval_in_days_average2 <- mean(merged_data$serial_interval_in_days2, na.rm = T)
serial_interval_in_days_sd2 <- sd(merged_data$serial_interval_in_days2, na.rm = T)

#Calculate serial interval in hours
merged_data$serial_interval_in_hours2 <- as.numeric(difftime(merged_data$infectiousness_end_infected, merged_data$infectiousness_end_infector, units = "hours"))

serial_interval_in_hours_average2 <- mean(merged_data$serial_interval_in_hours2, na.rm = T)
serial_interval_in_hours_sd2 <- sd(merged_data$serial_interval_in_hours2, na.rm = T)

print(serial_interval_in_hours_average2)
#28.9
print(serial_interval_in_hours_sd2)
#59.7
```

**Serial interval (By infectiousness start date)**

```{r}
#Calculate serial interval in hours
merged_data$serial_interval_in_hours3 <- as.numeric(difftime(merged_data$infectiousness_start_infected, merged_data$infectiousness_start_infector, units = "hours"))

serial_interval_in_hours_average3<- mean(merged_data$serial_interval_in_hours3, na.rm = T)
serial_interval_in_hours_sd3<- sd(merged_data$serial_interval_in_hours3, na.rm = T)

print(serial_interval_in_hours_average3)
#41.8
print(serial_interval_in_hours_sd3)
#46.5
```

## Generating Plots for Incubation Period

```{r}
#Filter incubation period
incubation_duration <- linelist$incubation_period[!(is.na(linelist$incubation_period)) & linelist$incubation_period > 0]

#Fit distribution
incubation_fit_ln_duration <- fitdist(incubation_duration, "lnorm")
summary(incubation_fit_ln_duration)

#Plot
denscomp(incubation_fit_ln_duration, main = "Density Plot of Incubation Period (Lognormal)") 
```

## Generating Plots for Illness Duration
```{r}
#Filter illness periods
illness_duration <- linelist$illness_duration[!(is.na(linelist$illness_duration)) & linelist$illness_duration > 0]

#Fit distribution
illness_fit_ln_duration <- fitdist(illness_duration, "lnorm")

summary(illness_fit_ln_duration)
#Plot
denscomp(illness_fit_ln_duration, main = "Density Plot of Illness Duration (Lognormal)")
```

## Generating Plots for Infectious Period
```{r}
#Filter Infectious periods
Infectious_duration <- linelist$infectious_period[!(is.na(linelist$infectious_period)) & linelist$infectious_period > 0]

#Fit distribution
Infectious_fit_ln_duration <- fitdist(Infectious_duration, "lnorm")

summary(Infectious_fit_ln_duration)

#Plot
denscomp(Infectious_fit_ln_duration, main = "Density Plot of Infectious Period (Lognormal)")
```

```{r, eval=FALSE}
merged_data$serial_interval_in_hours2
#Filter Infectious periods
serial_interval_in_hours <- merged_data$serial_interval_in_hours[!(is.na(merged_data$serial_interval_in_hours)) & merged_data$serial_interval_in_hours > 0]
#Fit distribution
Serial_interval_fit_ln_duration <- fitdist(serial_interval_in_hours, "lnorm")


summary(Serial_interval_fit_ln_duration)
#Plot
denscomp(Serial_interval_fit_ln_duration, main = "Density Plot of Serial Interval (Lognormal)")
```
Log Distribution of Natural History Parameters
```{r, eval=FALSE}
#Create Graph
par(mfrow=c(1,4))

#Incubation Period
incubation_duration <- linelist$incubation_period[!(is.na(linelist$incubation_period)) & linelist$incubation_period > 0]
incubation_fit_ln_duration <- fitdist(incubation_duration, "lnorm")

#Plot
denscomp(incubation_fit_ln_duration, main = "Incubation Period (hours)", xlab = "Incubation Period", ylab = "Density")

#Average
abline(v = mean(incubation_duration), col = "red", lty = 2)
text(mean(incubation_duration), 0.01, "Mean", col = "red")

#Illness Duration
illness_duration <- linelist$illness_duration[!(is.na(linelist$illness_duration)) & linelist$illness_duration > 0]
illness_fit_ln_duration <- fitdist(illness_duration, "lnorm")

#Plot
denscomp(illness_fit_ln_duration, main = "Illness Duration (hours)", xlab = "Illness Duration", ylab = "Density")

#Average
abline(v = mean(illness_duration), col = "red", lty = 2)
text(mean(illness_duration), 0.01, "Mean", col = "red")

#Infectious Period
infectious_duration <- linelist$infectious_period[!(is.na(linelist$infectious_period)) & linelist$infectious_period > 0]
infectious_fit_ln_duration <- fitdist(infectious_duration, "lnorm")

#Plot
denscomp(infectious_fit_ln_duration, main = "Infectious Period (hours)", xlab = "Infectious Period", ylab = "Density")

#Average
abline(v = mean(infectious_duration), col = "red", lty = 2)
text(mean(infectious_duration), 0.01, "Mean", col = "red")

#Serial Interval
serial_interval_in_hours <- merged_data$serial_interval_in_hours[!(is.na(merged_data$serial_interval_in_hours)) & merged_data$serial_interval_in_hours > 0]
serial_interval_fit_ln_duration <- fitdist(serial_interval_in_hours, "lnorm")

#Plot
denscomp(serial_interval_fit_ln_duration, main = "Serial Interval (hours)", xlab = "Serial Interval", ylab = "Density")

#Average
abline(v = mean(serial_interval_in_hours), col = "red", lty = 2)
text(mean(serial_interval_in_hours), 0.01, "Mean", col = "red")
```
Log Distribution of Natural History Parameters (Combined)
```{r, eval=FALSE}
#Create Graph
par(mfrow=c(1,1))

#Incubation Period
incubation_duration <- linelist$incubation_period[!(is.na(linelist$incubation_period)) & linelist$incubation_period > 0]
incubation_fit_ln_duration <- fitdist(incubation_duration, "lnorm")

#Illness Duration
illness_duration <- linelist$illness_duration[!(is.na(linelist$illness_duration)) & linelist$illness_duration > 0]
illness_fit_ln_duration <- fitdist(illness_duration, "lnorm")

#Infectious Period
infectious_duration <- linelist$infectious_period[!(is.na(linelist$infectious_period)) & linelist$infectious_period > 0]
infectious_fit_ln_duration <- fitdist(infectious_duration, "lnorm")

#Serial Interval
serial_interval_in_hours <- merged_data$serial_interval_in_hours[!(is.na(merged_data$serial_interval_in_hours)) & merged_data$serial_interval_in_hours > 0]
serial_interval_fit_ln_duration <- fitdist(serial_interval_in_hours, "lnorm")

#X-axis setting
max_duration <- max(c(max(incubation_duration), max(illness_duration), max(infectious_duration), max(serial_interval_in_hours)))

#Incubation Period
denscomp(incubation_fit_ln_duration, main = "Combined Density Plot", xlab = "Time (days)", ylab = "Density", xlim = c(0, max_duration))

#Illness Duration
lines(density(illness_duration), col = "blue", lty = 2, lwd = 2)  # Adjust the line thickness with lwd parameter

#Infectious Period
lines(density(infectious_duration), col = "green", lty = 2, lwd = 2)  # Adjust the line thickness with lwd parameter

#Serial Interval
lines(density(serial_interval_in_hours), col = "orange", lty = 2, lwd = 2)  # Adjust the line thickness with lwd parameter

#Legend
legend("topright", legend = c("Incubation", "Illness", "Infectious", "Serial Interval"), col = c("red", "blue", "green", "orange"), lty = 2, lwd = 2, cex = 0.8)

```



# Demographic Information

**Compare Period Prevalence of Undergraduates and Graduate Year 1 vs those that are Graduate Year 2 and #doctorate students**
```{r}
#Create new variable to dicotomize level variable
#Get numerator
full_data <- full_data %>% 
	mutate(education = ifelse(level %in% c("Graduate Year 1", "Senior"),0,1))

table(full_data$education)

# 44 = 0 and 37 = 1
#Get Denominator
roster <- roster %>% 
	mutate(education = ifelse(level %in% c("Graduate Year 1", "Senior"),0,1))

table(roster$education)
# 64 = 0 and 90 = 1

period_prevalance_G1S = 44/64
period_prevalance_G2D = 37/90
```


**Compare Period Prevalence of RFF for different PH programs**

```{r}
table(roster$program)

roster <- roster %>% 
  mutate(program_num = ifelse(program == "Epidemiology", 0,
                              ifelse(program == "Global Epidemiology", 1, 
                                     ifelse(program == "Global Health", 2, 3))))

table(roster$program_num)

full_data <- full_data %>% 
	mutate(program_num = ifelse(program == "Epidemiology", 0,
                              ifelse(program == "Global Epidemiology", 1, 
                                     ifelse(program == "Global Health", 2, 3))))
table(full_data$program_num)

period_prevalance_epi = 38/75
period_prevalance_glepi = 18/29
period_prevalance_gh = 12/29
period_prevalance_otherprs = 13/21
```
	
## Severity by Demographic Factors
```{r}
#Severity by Course
table(full_data$education)
table(Severe$education)

#Severity by Program
table(full_data$program_num)
table(Severe$program_num)

```

## Population Sizes
```{r}
table(roster$course)
#EPI_569 = 54, EPI 517 = 95, EPI 569 + EPI 517 = 5

table(roster$degree)
#Doctor of Philosphy = 4, Liberal Arts & Sciences = 1, Master of Public Health = 136, Master of Science Public Health = 10, Non-Degree Public Health = 3

table(roster$program)
#Behavioral Sciences = 3, Biostatistics = 1, Environmental Health + Epidemiology = 4, Environmental Health = 3, Epidemiology = 75, Global Environmental Health = 6, Global Epidemiology = 29, Global Health = 29, Human Health BA = 1, Special Standing Student = 3.

table(roster$level)
#full standing = 4, Graduate Year 1 = 63, Graduate Year 2 = 83, Non-Degree Graduate = 3, Senior = 1
```

## Case Distribution
```{r}
table(full_data$course.y)
#EPI 569 = 28 cases, EPI 517 = 49 cases, Both = 4

table(linelist$gender)
#Female = 49 cases, Male = 17 cases, Non-Binary = 2

table(Severe$gender)
#Females = 20 severe cases, Males = 8 severe cases
```

## Illness Duration in severe cases
```{r}
mean(Severe$illness_duration)
sd(Severe$illness_duration)
```

## Compare infectiousness between symptomatic and asymptomatic cases
```{r}
symptomatic <- linelist %>% 
	filter(symptomatic == 1)

asymptomatic <- linelist %>% 
	filter(symptomatic == 0)
#Symptomatic R0
mean(symptomatic$n_contacts)
sd(symptomatic$n_contacts)

#Asympomatic R0
mean(asymptomatic$n_contacts)
sd(asymptomatic$n_contacts)
```


# Reproduction Numbers
```{r}
start_day <- date("2023-10-11")
end_day <- date("2023-10-20")
```

We used the initial growth rate method to calculate $R_0$ from the plot of the log cases based on exposure date. The first case on RFF was observed on `r start_day`. Using the plot of log cases, the log-linear growth period ends on `r end_day`.
$I(t) \approx I(0) e^{rt}$

```{r}
linear_growth <- epicurve3 %>% 
	filter(exposure_date %in% c(start_day, end_day))
plot2 +
	geom_line(data = linear_growth,
						mapping = aes(x = exposure_date, y = log_cases),
						color = palette[2],
						linewidth = 1.5)
```

```{r}
r0_df <- epicurve3 %>%
	filter(exposure_date >= start_day, exposure_date <= end_day) %>% 
	splitstackshape::expandRows("incidence", drop = FALSE) %>% 
	dplyr::select(exposure_date) %>% 
	pull(exposure_date)

i <- incidence(r0_df)
mu <- serial_interval_in_days_average #replace with mean generation time
sigma <- serial_interval_in_days_sd #replace with standard deviation of generation time
res <- get_R(i, si_mean = mu, si_sd = sigma)
R_val <- sample_R(res, 1000)
r0 <- mean(R_val)
#2.41 (changes with each run)
```

Based on the initial growth rate, the basic reproductive number for RFF, $R_0$, is `r round(r0, 1)`.

We can compare this value to the $R_0$ calculated based purely on the number of people that each individual exposed.

```{r}
r0_2 <- mean(linelist$n_contacts, na.rm = TRUE)
#2.04

#By symptom status
r0_3 <- linelist %>% 
	group_by(symptomatic) %>% 
	summarize(r0 = mean(n_contacts, na.rm = TRUE),
						count = n())

#By severity
r0_4 <- linelist %>% 
	group_by(severe) %>% 
	summarize(r0 = mean(n_contacts, na.rm = TRUE),
						count = n())
```

# Modeling

## SEIR Equations

$\frac{dS}{dt} = -\lambda_t S$

$\frac{dE}{dt} = \lambda_t S - \gamma E$

$\frac{dI}{dt} = \gamma E - \sigma I$

$\frac{dR}{dt} = \sigma I$

## Waning Immunity Equations

$\frac{dS}{dt} = \omega R -\lambda_t S$

$\frac{dE}{dt} = \lambda_t S - \gamma E$

$\frac{dI}{dt} = \gamma E - 0.65 \sigma I - 0.35 \sigma I$

$\frac{dR}{dt} = 0.65 \sigma I - \omega R$

$\frac{dR_s}{dt} = 0.35 \sigma I$

## Observed Data

```{r}
mean_duration_severe <- linelist %>%
	filter(severe == 1) %>%
	summarize(mean(illness_duration, na.rm = TRUE)) %>% 
	as.numeric()
#24.39286

mean_duration_nonsevere <- linelist %>%
	filter(severe == 0) %>%
	summarize(mean = mean(illness_duration, na.rm = TRUE)) %>% 
	as.numeric()
#25.66667

mean_latent <- mean(linelist$latent_period, na.rm = TRUE)
#23.36364

```

```{r}
#Create function for model equations
sir_ode <- function(times,init,parms){
  with(as.list(c(parms,init)), {
    # ODEs
    dS <- - (alpha*beta*I*S/(S+E+I+R))
    dE <- (alpha*beta*I*S/(S+E+I+R)) - (gamma*E)
    dI <- (gamma*E) - (sigma*I)
    dR <- (sigma*I)
    list(c(dS,dE,dI,dR))
  })
}

#Define parameters
parms <- c(alpha = 5,              # alpha = daily contacts
           beta = 0.28,             # beta = probability of infection on contact
				 	 gamma = 1/(mean_latent/24),  # gamma = rate of infectiousness onset
           sigma = 1/(mean_duration/24))# sigma = rate of recovery per day

#Set initial conditions
init <- c(S = 153,
					E = 0,
					I = 1,
					R = 0)

times <- seq(0, 30, length.out=30)
sir_out <- lsoda(init,times,sir_ode,parms)
sir_out_long <- melt(as.data.frame(sir_out),"time")

#Plot the model output
sir_out_long %>% ggplot(aes(x=time, y=value, colour=variable, group=variable)) +
	geom_line(lwd = 2) +
  labs(x = "Time since index case exposure (days)",
  		 y = "Number of cases",
  		 title = "SEIR Model of RFF") +
	scale_color_manual(name = "Class",
										 breaks = c("S", "E", "I", "R"),
										 labels = c("Susceptible",
										 					"Exposed",
										 					"Infectious",
										 					"Recovered"),
										 values = c("S" = as.character(palette["blue"]),
										 					 "E" = as.character(palette["dark_green"]),
										 					 "I" = as.character(palette["green"]),
										 					 "R" = as.character(palette["red"]))) +
	theme(panel.grid = element_blank())

as.numeric(sir_out_long %>% 
	filter(variable == "R") %>% 
	summarize(max(value)))
```

## Waning Immunity

```{r}
mean_duration_severe <- linelist %>%
	filter(severe == 1) %>%
	summarize(mean = mean(illness_duration, na.rm = TRUE)) %>% 
	as.numeric()
#24.39286

mean_duration_nonsevere <- linelist %>%
	filter(severe == 0) %>%
	summarize(mean = mean(illness_duration, na.rm = TRUE)) %>% 
	as.numeric()
#25.66667

mean_latent <- as.numeric(mean_latent)
#23.36364
```

```{r}
#Create function for model equations
sir_ode <- function(times,init,parms){
  with(as.list(c(parms,init)), {
    # ODEs
    dS <- (omega*R) - (S*alpha*beta*I/(S+E+I+R+R_s))
    dE <- (S*alpha*beta*I/(S+E+I+R+R_s)) - (gamma*E)
    dI <- (gamma*E) - (0.65*sigma*I) - (0.35*mu*I)
    dR <- (0.65*sigma*I)  - (omega*R)
    dRs <- 0.35*mu*I
    list(c(dS,dE,dI,dR,dRs))
  })
}

#Define parameters
parms <- c(alpha = 5,              # alpha = daily contacts
           beta = 0.62,             # beta = probability of infection on contact
				 	 gamma = 1/(mean_latent/24),  # gamma = rate of infectiousness onset
           sigma = 1/(mean_duration/24),# sigma = rate of recovery per day
           mu = 1/(mean_duration/24),   # mu = rate of severe disease
           omega = 0)            # omega = rate of immune loss per day

#Set initial conditions
init <- c(S = 153,
					E = 0,
					I = 1,
					R = 0,
					R_s = 0)

times <- seq(0, 30, length.out=30)
sir_out <- lsoda(init,times,sir_ode,parms)
sir_out_long <- melt(as.data.frame(sir_out),"time")

#Plot the model output
sir_out_long %>% ggplot(aes(x=time, y=value, colour=variable, group=variable)) +
	geom_line(lwd = 2) +
  labs(x = "Time since index case exposure (days)",
  		 y = "Number of cases",
  		 title = "SEIR Model of RFF") +
	scale_color_manual(name = "Class",
										 breaks = c("S", "E", "I", "R", "R_s"),
										 labels = c("Susceptible",
										 					"Exposed",
										 					"Infectious",
										 					"Recovered",
										 					"Became severe"),
										 values = c("S" = as.character(palette["blue"]),
										 					 "E" = as.character(palette["dark_green"]),
										 					 "I" = as.character(palette["green"]),
										 					 "R" = as.character(palette["red"]),
										 					 "R_s" = as.character(palette["dark_red"]))) +
	theme(panel.grid = element_blank())

as.numeric(sir_out_long %>% 
	filter(variable == "R") %>% 
	summarize(max(value))) +
	as.numeric(sir_out_long %>% 
	filter(variable == "R_s") %>% 
	summarize(max(value)))
```

# Time-varying Reproductive Number (Rt)

```{r}
#Convert data to a format that can be used in the EpiEstim wallinga_teunis function 
## Note: to calculate an Rt estimate for day 1 of the outbreak, we must start our epi curve 2 days prior the first symptom onset date

epicurve5 <- epicurve2 %>%
	arrange(exposure_date) %>% 
	rename(date = exposure_date, I = incidence) %>% 
	dplyr::select(date, I)

all.dates <- as.data.frame(seq(as.Date("2023-10-09"), by = "day", length.out = 23)) 
names(all.dates) <- "date"

epicurve.epiestim <- merge(x=epicurve5, y=all.dates, by="date", all="TRUE")
epicurve.epiestim <- epicurve.epiestim %>% mutate(I = ifelse(is.na(I), 0, I)) 

#Estimate Rt, along with 95% confidence intervals for Rt estimates

mean_si_days <- 41.12/24   
std_si_days <- 47.15/24

estimates <- wallinga_teunis(epicurve.epiestim$I, 
                             method="parametric_si",
                             config = list(t_start = seq(1, 23), 
                                           t_end = seq(1, 23),
                                           mean_si = mean_si_days, 
                                           std_si = std_si_days, 
                                           n_sim = 1000))

#save estimates R to a new dataframe 
#set zeros for where estimates are missing 

estimates_R <- as.data.frame(estimates$R)
	
	
estimates_R <- as.data.frame(estimates$R) %>%
	rename(mean = `Mean(R)`,
				 std = `Std(R)`,
				 Q25 = `Quantile.0.025(R)`,
				 Q975 = `Quantile.0.975(R)`) %>% 
	rbind(data.frame(t_start = c(6, 7, 21),
									 t_end = c(6, 7, 21),
									 mean = 0,
									 std = 0,
									 Q25 = 0,
									 Q975 = 0)) %>%
	arrange(t_start)

#Examine the serial interval distribution
plot(estimates$si_distr, xlab="Serial Interval (Days)", ylab="Proportion")

```

Plot the R(t) estimates and 95% CIs over the epi curve

```{r}
plot2.data <- cbind(epicurve.epiestim[-(1:4),], estimates_R$mean,
                    estimates_R$Q25, estimates_R$Q975)
names(plot2.data) <- c("dates", "I", "R", "lowerCI", "upperCI")

```

```{r}
plot2 <- ggplot(data=plot2.data, aes(x=dates, y=I, labs=FALSE)) + 
  geom_bar(stat="identity", fill="#636467") +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous(breaks = seq(0,20, by = 2)) +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(x = "Exposure Date", 
       y = "Number of Cases (bars) and Rt (line;95% CI)",
       title = "Epi Curve for RFF Outbreak (Mean SI=1.7)") +
  geom_hline(aes(yintercept=1), colour="#E28571", linetype="dashed", size=0.5) +
  geom_errorbar(data=plot2.data, aes(ymax=upperCI, ymin=lowerCI, width=0.6),stat="identity", size=0.8, show.legend=FALSE) +
  geom_line(data=plot2.data[!is.na(plot2.data$R),],aes(x=dates, y=R), color='#66B4BE', size=0.5) +
  geom_point(data = plot2.data, aes(x=dates, y=R), size=1.2, show.legend=FALSE) 
plot2
```

# Calculate critical vaccination thresholds

$P_c = (1 - 1/R_0) / VE$ 

$HIT = (1 - 1/R_0)$ 

$P_c = HIT / VE$ 

$R_0 = 2.41$ 

```{r}

R0 <- 2.41 
HIT <- 1 - (1/R0)
HIT

#VE = 100%
Pc_100 <- (HIT/1) *100
Pc_100

#VE = 90% 
Pc_90 <- (HIT/0.9) *100
Pc_90

#VE = 80% 
Pc_80 <- (HIT/0.8) *100
Pc_80

#VE = 70%
Pc_70 <- (HIT/0.7) *100
Pc_70

#VE = 60%
Pc_60 <- (HIT/0.6) *100
Pc_60

#VE = 50%
Pc_50 <- (HIT/0.5) *100
Pc_50

#VE = 50%
Pc_40 <- (HIT/0.4) *100
Pc_40

```

