---
title: "EPI 569 Final Project"
author: "Meghna Ray"
date: "`r Sys.Date()`"
output: html_document
---

### R setup 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***Load packages***
```{r}
pacman::p_load(tidyverse, 
               ggplot2,
               janitor,
               lubridate,
               dplyr)
```

### Load data
```{r}
raw_data <- readRDS('rff_finalcases_clean_2023.rds') %>% janitor::clean_names()
roster <- readRDS('rosters_569_517_clean_2023.rds') %>%  janitor::clean_names()
```

### Data Cleaning 
```{r}
linelist <- raw_data %>% 
  #rename variables 
  rename(c('report' = 'reporting_time',
           'infector_id' = 'infectedby',
           'exposure_date' = 'date_of_exposure',
           'exposure_time' = 'time_of_exposure',
           'symptomatic' = 'did_you_have_symptoms',
           'incubation_period' = 'how_many_hours_after_exposure_did_you_develop_symptoms',
           'illness_duration' = 'how_many_hours_after_your_symptom_onset_did_you_feel_better',
           'n_contacts' = 'how_many_people_did_you_expose',
           'n_infect' = 'how_many_people_did_you_actually_infect',
           'contact1_id' = 'case_id_first_exposure',
           'contact1_date' = 'date_of_first_exposure',
           'contact1_time' = 'time_of_first_exposure',
           'contact2_id' = 'case_id_second_exposure',
           'contact2_date' = 'date_of_second_exposure',
           'contact2_time' = 'time_of_second_exposure',
           'contact3_id' = 'case_id_third_exposure',
           'contact3_date' = 'date_of_third_exposure',
           'contact3_time' = 'time_of_third_exposure',
           'contact4_id' = 'case_id_fourth_exposure',
           'contact4_date' = 'date_of_fourth_exposure',
           'contact4_time' = 'time_of_fourth_exposure',
           'contact5_id' = 'case_id_fifth_exposure',
           'contact5_date' = 'date_of_fifth_exposure',
           'contact5_time' = 'time_of_fifth_exposure',
           'exposure0' = 'date_time_exposure',
           'exposure1' = 'date_time_first_exposure',
           'exposure2' = 'date_time_second_exposure',
           'exposure3' = 'date_time_third_exposure',
           'exposure4' = 'date_time_fourth_exposure',
           'exposure5' = 'date_time_fifth_exposure')) %>% 
  
  #recode gender variable as numeric (0 = Female, 1 = Male, 2 = Non-binary)
  mutate( gender = recode(gender, "Female" = 0, "Male" = 1, "Non-binary" = 2)) %>% 
  
  #recode course variable as numeric (0 = EPI 569, 1 = EPI 517, 2 = both)
  mutate(course = recode(course, "EPI 569 (Concepts and Methods in ID EPI)" = 0, 
                         "EPI/GH 517 (Case Studies in ID)" = 1,
                         "BOTH" = 2)) %>% 
  
  #recode symptomatic variable as numeric (0 = No, 1 = Yes)
  mutate(symptomatic = recode(symptomatic, "No" = 0, "Yes" = 1))
  
```


### Plot Epi Curve

***Onset variable***
```{r}
#Create new onset variable using date and time of exposure + incubation period 
linelist <- linelist %>% 
  mutate(onset = exposure0 + incubation_period) %>% 
  mutate(onset = date(onset))
```

***Plot epi curve***
```{r}
#Create data frame for epi curve with onset and incidence
epicurve <- as.data.frame(linelist %>% 
                            group_by(onset) %>%
                            summarize(incidence=n()))

#Plot epi curve 
plot1 <- epicurve %>% 
  ggplot(aes(x = onset, y = incidence, labs = FALSE)) + 
  geom_bar(stat="identity", fill="gray45") +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous(breaks = seq(0, 15, by = 2)) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Onset Date",
       y = "Number of Cases",
       title = "Epi Curve for Rollins Fall Fever")
plot1
```

***Plot log cases***
```{r, include = TRUE, echo = T}
epicurve <- epicurve %>% 
  mutate(log_cases = log(incidence))

#Plot log cases
plot2 <- epicurve %>% 
  ggplot(aes(x = onset, y = log_cases, labs = FALSE)) + 
  geom_bar(stat="identity", fill="gray45") +
  scale_x_date(date_breaks = "2 days", date_labels = "%m/%d") +
  scale_y_continuous() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Exposure Date",
       y = "Log Cases",
       title = "Epi Curve - Log Cases")
plot2
```

***Recovered variable***
```{r}
#Create new onset variable using date and time of exposure + incubation period 
linelist <- linelist %>% 
  mutate(recovered = onset + illness_duration)
```

***Calculate number of recovered by day***
```{r}
recovered <- linelist %>% 
  group_by(recovered) %>% 
  summarize(n_recovered = n()) %>% 
  na.omit()
```

***Calculate number of infected by day***
```{r}
infected <- linelist %>% 
  group_by(onset) %>% 
  summarize(n_onset = n()) %>% 
  na.omit()

#option 2
dateseq <- seq(from = as.Date(min(linelist$onset, na.rm = TRUE)),
								 to = as.Date(max(linelist$recovered, na.rm = TRUE)),
								 by = "1 day")

infectedseq <- map(dateseq, ~ sum(.x >= linelist$onset & .x < linelist$recovered, na.rm = TRUE))

infected2$date <- dateseq
infected2$n_infected <- as.numeric(infectedseq)

#I tried this instead and got infected2 to work (Bevin)
infectedseq <- map(dateseq, ~sum(.x >= linelist$onset & .x < linelist$recovered, na.rm = TRUE))
infected2 <- data.frame(date = dateseq, n_infected = as.numeric(infectedseq))
```

***Calculate number of susceptible by day***
```{r}

#Join infected and recovered datasets
epicurve2 <- infected2 %>% 
  full_join(recovered, by =c('date' = 'recovered')) %>% 
	full_join(infected, by = c("date" = "onset"))

#Set na values to zero 
epicurve2 <- epicurve2 %>% 
  mutate(n_infected = ifelse(is.na(n_infected), 0, n_infected)) %>% 
  mutate(n_recovered = ifelse(is.na(n_recovered), 0, n_recovered)) %>% 
	mutate(n_onset = ifelse(is.na(n_onset), 0, n_onset))

#Create variable for cumulative recoveries and infected
epicurve2 <- epicurve2 %>% 
  mutate(total_recovered = cumsum(n_recovered)) %>% 
  mutate(total_infected = cumsum(n_onset))

#Calculate number of people not susceptible and susceptible 
total_pop <- nrow(roster)
epicurve2 <- epicurve2 %>% 
  mutate(susceptible = total_pop - total_infected)
```

***Plot susceptible, infected, and recovered by day***
```{r}
plot2 <- ggplot(epicurve2, aes(x = date)) +
  geom_line(aes(y = susceptible, color = "Susceptible")) +
  geom_line(aes(y = n_infected, color = "Infected")) +
  geom_line(aes(y = total_recovered, color = "Recovered")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Date",
       y = "Number of Cases",
       title = "Epi Curve for Rollins Fall Fever") +
  scale_color_manual(name='Class',
                     breaks=c('Susceptible', 'Infected', 'Recovered'),
                     values=c('Susceptible'='red', 'Infected'='blue', 'Recovered'='green')) +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 8))

plot2
```

***Wanted to see how a percent Epi Curve would look. I havent seen the data presented this way in class though***
```{r}
epicurve3 <- epicurve2 %>% 
	mutate(susceptible2 = susceptible/153) %>% 
	mutate(n_infected2 = n_infected/153) %>% 
	mutate(total_recovered2 = total_recovered/153)

plot3 <- ggplot(epicurve3, aes(x = date)) +
  geom_line(aes(y = susceptible2, color = "Susceptible")) +
  geom_line(aes(y = n_infected2, color = "Infected")) +
  geom_line(aes(y = total_recovered2, color = "Recovered")) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Date",
       y = "Proportion of Total Population",
       title = "Epi Curve for Rollins Fall Fever") +
  scale_color_manual(name='Class',
                     breaks=c('Susceptible', 'Infected', 'Recovered'),
                     values=c('Susceptible'='red', 'Infected'='blue', 'Recovered'='green')) +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 8))

plot3
```

