---
title: "Natural History"
output: html_document
date: "2023-11-24"
---

***Load packages***
```{r}
pacman::p_load(tidyverse, 
							 ggplot2,
							 janitor,
							 lubridate,
							 dplyr,
							 fitdistrplus)
```

### Load data
```{r}
raw_data <- readRDS('/Users/bevinmanuelpillai/Desktop/EPI 569/RFF/rff_finalcases_clean_2023.rds') %>% janitor::clean_names()
roster <- readRDS('/Users/bevinmanuelpillai/Desktop/EPI 569/RFF/rosters_569_517_clean_2023.rds') %>%  janitor::clean_names()
```

### Data Cleaning 
```{r}
linelist <- raw_data %>% 
	#rename variables 
	rename(c('report' = 'reporting_time',
					 'infector_id' = 'infectedby',
					 'exposure_date' = 'date_of_exposure',
					 'exposure_time' = 'time_of_exposure',
					 'symptomatic' = 'did_you_have_symptoms',
					 'incubation_period' = 'how_many_hours_after_exposure_did_you_develop_symptoms',
					 'illness_duration' = 'how_many_hours_after_your_symptom_onset_did_you_feel_better',
					 'n_contacts' = 'how_many_people_did_you_expose',
					 'n_infect' = 'how_many_people_did_you_actually_infect',
					 'contact1_id' = 'case_id_first_exposure',
					 'contact1_date' = 'date_of_first_exposure',
					 'contact1_time' = 'time_of_first_exposure',
					 'contact2_id' = 'case_id_second_exposure',
					 'contact2_date' = 'date_of_second_exposure',
					 'contact2_time' = 'time_of_second_exposure',
					 'contact3_id' = 'case_id_third_exposure',
					 'contact3_date' = 'date_of_third_exposure',
					 'contact3_time' = 'time_of_third_exposure',
					 'contact4_id' = 'case_id_fourth_exposure',
					 'contact4_date' = 'date_of_fourth_exposure',
					 'contact4_time' = 'time_of_fourth_exposure',
					 'contact5_id' = 'case_id_fifth_exposure',
					 'contact5_date' = 'date_of_fifth_exposure',
					 'contact5_time' = 'time_of_fifth_exposure',
					 'exposure0' = 'date_time_exposure',
					 'exposure1' = 'date_time_first_exposure',
					 'exposure2' = 'date_time_second_exposure',
					 'exposure3' = 'date_time_third_exposure',
					 'exposure4' = 'date_time_fourth_exposure',
					 'exposure5' = 'date_time_fifth_exposure')) %>% 
	
	#recode gender variable as numeric (0 = Female, 1 = Male, 2 = Non-binary)
	mutate( gender = recode(gender, "Female" = 0, "Male" = 1, "Non-binary" = 2)) %>% 
	
	#recode course variable as numeric (0 = EPI 569, 1 = EPI 517, 2 = both)
	mutate(course = recode(course, "EPI 569 (Concepts and Methods in ID EPI)" = 0, 
												 "EPI/GH 517 (Case Studies in ID)" = 1,
												 "BOTH" = 2)) %>% 
	
	#recode symptomatic variable as numeric (0 = No, 1 = Yes)
	mutate(symptomatic = recode(symptomatic, "No" = 0, "Yes" = 1))

```


### Create Onset Variable
```{r}
#Create new onset variable using date and time of exposure + incubation period 
linelist <- linelist %>% 
	mutate(onset = exposure0 + incubation_period) %>% 
	mutate(onset = date(onset))
```

### Calculating mean incubation period
```{r}
linelist %>%
	summarise(mean(incubation_period, na.rm = T))
#23.36
```

### Calculating median incubation period
```{r}
linelist %>%
	summarise(median(incubation_period, na.rm = T))
#24
```

### Calculating mean ill duration
```{r}
linelist %>%
	summarise(mean(illness_duration, na.rm = T))
#25.02
```


### Calculating median ill duration
```{r}
linelist %>%
	summarise(median(illness_duration, na.rm = T))
#26
```
Serial Interval
```{r}
serial <- raw_data[, c("case_id", "infectedby", "date_of_exposure", "time_of_exposure", "how_many_hours_after_exposure_did_you_develop_symptoms")]

#Covert to Exposure to Date_Time format
serial$date_time_exposure <- as.POSIXct(paste(serial$date_of_exposure, serial$time_of_exposure), format="%Y-%m-%d %H:%M:%S")
serial <- serial[, !(names(serial) %in% c("date_of_exposure", "time_of_exposure"))]

#Covert Symptom Onset to Date_Time Format
serial$date_time_symptom_onset <- serial$date_time_exposure + 
  as.difftime(serial$how_many_hours_after_exposure_did_you_develop_symptoms, units = "hours")

serial <- serial[, !(names(serial) %in% c("how_many_hours_after_exposure_did_you_develop_symptoms"))]

#Get infector and infected columns
merged_data <- merge(serial, serial, by.x = "infectedby", by.y = "case_id", suffixes = c("_infected", "_infector"))

#Calculate serial interval
merged_data$serial_interval_in_days <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "days"))

merged_data$serial_interval_in_hours <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "hours"))

merged_data$serial_interval_in_mins <- as.numeric(difftime(merged_data$date_time_symptom_onset_infected, merged_data$date_time_symptom_onset_infector, units = "mins"))

mean(merged_data$serial_interval_in_mins, na.rm = T)
median(merged_data$serial_interval_in_mins, na.rm = T)

mean(merged_data$serial_interval_in_days, na.rm = T)


```
#Calculating Mean Latent Period
```{r}
serial$latent <- serial$date_time_symptom_onset - serial$date_time_exposure
mean_latent <- mean(serial$latent, na.rm = TRUE)
print(mean_latent)
```


#Generating Plots for Incubation Period
```{r}
# Filter incubation periods
incubation_duration <- linelist$incubation_period[!(is.na(linelist$incubation_period)) & linelist$incubation_period > 0]

# Fit a lognormal distribution
incubation_fit_ln_duration <- fitdist(incubation_duration, "lnorm")

# Summarize the fitted distribution
summary(incubation_fit_ln_duration)

denscomp(incubation_fit_ln_duration)
```
#Generating Plots for illness_duration
```{r}
# Filter incubation periods
illness_duration <- linelist$illness_duration[!(is.na(linelist$illness_duration)) & linelist$illness_duration > 0]

# Fit a lognormal distribution
illness_fit_ln_duration <- fitdist(illness_duration, "lnorm")

# Summarize the fitted distribution
summary(illness_fit_ln_duration)

denscomp(illness_fit_ln_duration)
```

#Calculating infectious period for each person (assuming exposure is start of infectiousness)
```{r}
#Covert to exposure times to Date_Time format
linelist$exposure_date_time <- as.POSIXct(paste(linelist$exposure_date, linelist$exposure_time), format="%Y-%m-%d %H:%M:%S")
linelist <- linelist[, !(names(linelist) %in% c("exposure_date", "exposure_time"))]

#Covert to contact times to Date_Time format
linelist$contact1_date_time <- as.POSIXct(paste(linelist$contact1_date, linelist$contact1_time), format="%Y-%m-%d %H:%M:%S")
linelist <- linelist[, !(names(linelist) %in% c("contact1_date", "contact1_time"))]

linelist$contact2_date_time <- as.POSIXct(paste(linelist$contact2_date, linelist$contact2_time), format="%Y-%m-%d %H:%M:%S")
linelist <- linelist[, !(names(linelist) %in% c("contact2_date", "contact2_time"))]


linelist$contact3_date_time <- as.POSIXct(paste(linelist$contact3_date, linelist$contact3_time), format="%Y-%m-%d %H:%M:%S")
linelist <- linelist[, !(names(linelist) %in% c("contact3_date", "contact3_time"))]


linelist$contact4_date_time <- as.POSIXct(paste(linelist$contact4_date, linelist$contact4_time), format="%Y-%m-%d %H:%M:%S")
linelist <- linelist[, !(names(linelist) %in% c("contact4_date", "contact4_time"))]


linelist$contact5_date_time <- as.POSIXct(paste(linelist$contact5_date, linelist$contact5_time), format="%Y-%m-%d %H:%M:%S")
linelist <- linelist[, !(names(linelist) %in% c("contact5_date", "contact5_time"))]


#Calculate time between exposure and last contact
linelist$time_between_exposure_last_contact <- with(linelist, ifelse(!is.na(contact5_date_time), contact5_date_time - exposure_date_time,
                                                                     ifelse(!is.na(contact4_date_time), contact4_date_time - exposure_date_time,
                                                                     ifelse(!is.na(contact3_date_time), contact3_date_time - exposure_date_time,
                                                                     ifelse(!is.na(contact2_date_time), contact2_date_time - exposure_date_time,
                                                                     ifelse(!is.na(contact1_date_time), contact1_date_time - exposure_date_time, NA))))))


#Calculate mean infectious period based on start of exposure and last contact
mean_time <- linelist %>% 
  filter(case_id != 6) %>% 
  summarise(mean = mean(time_between_exposure_last_contact, na.rm = T))

#1463.86
```

#Calculating infectious period for each person (assuming symptom_onset is start of infectiousness)
```{r}
d <- left_join(linelist, serial, by = "case_id")

#Calculate time between symtpom and last contact
d$time_between_symptoms_last_contact <- with(d, ifelse(!is.na(contact5_date_time), contact5_date_time - date_time_symptom_onset,
                                                                     ifelse(!is.na(contact4_date_time), contact4_date_time - date_time_symptom_onset,
                                                                     ifelse(!is.na(contact3_date_time), contact3_date_time - date_time_symptom_onset,
                                                                     ifelse(!is.na(contact2_date_time), contact2_date_time - date_time_symptom_onset,
                                                                     ifelse(!is.na(contact1_date_time), contact1_date_time - date_time_symptom_onset, NA))))))


#Calculate mean infectious period based on start of exposure and last contact
mean_time <- d %>% 
  filter(case_id != 6) %>% 
  summarise(mean = mean(time_between_symptoms_last_contact, na.rm = T))
```


```{r}
d2 <- d[, c("case_id","infectedby","exposure0", "exposure1", "exposure2", "exposure3", "exposure4", "exposure5","date_time_symptom_onset")]
# Assuming 'd2' is your data frame
date_columns <- c("exposure0", "exposure1", "exposure2", "exposure3", "exposure4", "exposure5", "date_time_symptom_onset")

d2[date_columns] <- lapply(d2[date_columns], function(x) as.POSIXct(x, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"))

# Create a new variable 'outlier'
d2$outlier <- ifelse(d2$date_time_symptom_onset < do.call(pmin, c(d2[, c("exposure1", "exposure2", "exposure3", "exposure4", "exposure5")], na.rm = TRUE)), 0, 1)

d3 <- d2 %>% 
	filter(outlier == 1)

outlier_vector <- d3$case_id

print(outlier_vector)
```
Infectious Period(symptom onset to symptom recovery)
```{r}
#Convert exposure to Date_Time Format
infectious <- raw_data %>% 
  mutate(date_time_exposure = as.POSIXct(paste(date_of_exposure, time_of_exposure), format="%Y-%m-%d %H:%M:%S"))

infectious <- infectious[, !(names(infectious) %in% c("date_of_exposure", "time_of_exposure"))]
	
#Calculate Symptom Onset Date_Time
#Covert Symptom Onset to Date_Time Format
infectious$date_time_symptom_onset <- infectious$date_time_exposure + 
  as.difftime(infectious$how_many_hours_after_exposure_did_you_develop_symptoms, units = "hours")

infectious <- infectious[, !(names(infectious) %in% c("how_many_hours_after_exposure_did_you_develop_symptoms"))]
#Calculate Recovery Date_time
infectious$date_time_recovery <- infectious$date_time_symptom_onset + 
  as.difftime(infectious$how_many_hours_after_your_symptom_onset_did_you_feel_better, units = "hours")

infectious <- infectious[, !(names(infectious) %in% c("how_many_hours_after_your_symptom_onset_did_you_feel_better"))]

infectious$infectious_period <- infectious$date_time_recovery -infectious$date_time_symptom_onset
	
```

